---
output: 
  stevetemplates::article:
    fig_caption: true
#bibliography: master.bib
biblio-style: apsr
title: "Recreating David Card's \"Using Geographic Variation In College Proximity To Estimate The Return To Schooling\" 1993 Working Paper"
#thanks: "Replication files are available on the author's Github account (http://github.com/svmiller). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: svmille@clemson.edu."
author:
- name: Shane Ellard
  affiliation: York University
abstract: "In this short paper I recreate the main analysis done by David Card in his 1993 working paper \"Using Geographic Variation In College Proximity To Estimate The Return To Schooling\". "
#keywords: "pandoc, r markdown, knitr"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=2in
fontfamily: mathpazo
fontsize: 11
spacing: double
endnote: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      echo = FALSE,
                      message=FALSE, warning=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })

# kable
options(knitr.kable.NA = '')

# Set the graphical theme
ggplot2::theme_set(ggplot2::theme_light())

library(tidyverse)
library(AER)
library(fixest)
library(haven)
library(kableExtra)
library(modelsummary)
```


# The Economic Returns to Schooling

One persistent fact of labour markets is that better-educated workers earn more. Despite this, a comparison of the earnings between high school and college educated workers, for example, would yield a biased estimate of the return to schooling. Education levels are not randomly assigned. Rather, individuals make decisions about how much schooling they receive. If these decisions about education are correlated with an unobserved component of earnings then a difference in means comparison would over-state or under-state the true return to education. In this case, education is said to be endogenous in the earnings equation.

A standard way to address this issue is through the use of an instrumental variable. An instrument is some observable variable that effects the schooling decision but \underline{does not} directly effect earnings. If we have a valid instrument, then the variation in education induced by the instrument is uncorrelated with the error term in the earnings equation. It is this variation we exploit in IV regression. In general, valid instruments are difficult to find. In his 1993 working paper, Card proposes using proximity to college as an instrument for education. It is easy to see why college proximity would effect the schooling decision. Living close to college gives students the option of living at home and thus lowers the cost of attending school, at least among children from low-income families. If college proximity also does not directly enter the earnings equation then it can be used as an instrument. 

With this in mind, Card drew data from the National Longitudinal Survey of Young Men (NLSYM)\footnote{Card has provided these data and they are freely available through the Wooldridge data package.}. Card ran a series of basic OLS and IV specifications, using proximity to college as an instrument to regress log wages on education. He found that among these men when college proximity is used as an instrument for education the implied returns are about 50% higher then the respective OLS estimates. In the coming sections I will replicate Card's main results; including his OLS specifications, IV specifications, and his tests to evaluate if college proximity is a legitimate instrument.

# Returns to Schooling using OLS

To begin his analysis Card considered five OLS estimates for the returns to schooling. Table 1 presents these same specifications. All five estimations include a linear education term, a quadratic function of experience, a race indicator, and dummy variables for residence in the south and in a metropolitan area (SMSA). The second adds indicator variables for the region of residence in 1966 and another for residence in a SMSA in 1966. The third estimation adds measures of the mother's and father's education, and the fourth adds an interaction between these education terms. The fifth estimation adds indicator variables for family structure. 

All five of these estimations yield similar results for the returns on schooling. This can be quite easily seen by looking at the coefficients in the first row of Table 1. Regardless of the controls used, the basic OLS estimates imply a return to education of about 7.3 percent per year of schooling. 

Despite similar estimates across specifications one should still be skeptical of these results. Recall, we suspect that education is correlated with an unobserved part of earnings and that our estimate is therefore biased. There is a range of reasons why we are worried this might be the case. One reason is the so-called "ability bias". If individuals have an unobserved characteristic "ability" that raises wages and also influences one to receive more schooling then the OLS estimate of the return to schooling will be biased upwards. Another potential issue that Card mentions is measurement error. Measurement error induces a negative correlation between the errors in earnings and schooling, this creates a downward bias in the OLS estimate of the return to schooling. Similarly, a downward bias occurs if true returns to schooling are heterogeneous and if those with lower levels of education have higher returns to schooling\footnote{For a detailed discussion of these problems see Card (1999)}.

# Returns to Schooling using IV

Table 2 replicates the basic IV estimates of the return to education done by Card, using college proximity as an instrument for completed education. Specifically, a variable indicating whether one is near an accredited 4-year college is used as the instrument. Columns one and two report the coefficients for the first stage regressions (the effect of college proximity on years of schooling). Columns three and four contain the reduced form regressions (the effect of college proximity on log wages). Columns five and six then report IV estimates of the return of schooling. The regressions in columns one, three and five exclude parental education and family structure controls whereas columns two, four and six include these controls.

As Card addressed, one potential problem is that if education is endogenous to the earnings equation then so is experience, as it is mechanically related to education. To address this, Card replicated the same IV specifications as above but used age as an instrument for experience. These specifications are replicated in Table 3. Like before, columns one, three and five exclude familial controls whereas columns two, four and six include them.

The implementations yield similar results regardless what controls are included or how experience is treated. Living near college had a strong positive effect on schooling (between .32 and .38 years) and a positive effect on wages (between 4.2 and 4.8 percent). Using proximity to college as an instrument for education then yields estimates of then returns to schooling between 12.2 and 13.6 percent. Like Card found, the IV estimates on the return to schooling are about 50 percent higher then the corresponding OLS estimates.

To probe the robustness of these findings Card implemented several alternative specifications. Tables 4 and 5 replicates some of these specifications. Table 4 contains the alternative specifications using OLS and Table 5 contains the alternative IV specifications. The first columns of these tables contain the basic OLS and IV specifications from above. All of the controls for experience, race, region and family background are included, and age is used as an instrument for experience. The second columns then control for "ability" as directly measured by the "Knowledge of World Work" (KWW) score. Like Card found, in the OLS estimates KWW is a significant predictor of earnings and the inclusion of KWW also leads to a reduction in the estimated returns to schooling. However, when it is included in the IV specification the coefficient on KWW is small and statistically insignificant. 

A criticism of this specification is that KWW is treated as an error-free measure of ability. To address this, Card used IQ to instrument KWW. This specification is replicated in the third columns of Tables 4 and 5. As Card found, doing this lowers the IV estimate of the return to schooling but increased the standard errors of the coefficients of education and KWW score to where neither is statistically different from zero.

The final column in Table 4 uses an alternative measure of college proximity to instrument education. Specifically, indicators for proximity to a 4-year college and proximity to a 2-year college are both used as instruments. Using both indicators as instruments leads to a small reduction in the estimated returns to schooling, but nevertheless the implied return to education and standard errors are very similar to the previous IV estimations. 

Altogether, the results in Tables 4 and 5 confirm our findings from the initial IV specifications. That is, the returns to schooling estimated when using college proximity as an instrument are uniformly higher than the OLS estimates. Moreover, regardless of what controls are used, the IV estimates on the returns to schooling are about 40-50% greater then the respective OLS estimates. 

# Is College Proximity a Valid Instrument?

As Card notes, for college proximity to serve as a valid instrument for education it must have no direct effect on earnings (that is, it is exogenous to earnings). There are several reasons that Card discusses for why this might not be the case. One possibility is that families who place a large emphasis on education will choose to live closer to colleges. Their children might also have higher "ability" or may be more motivated to achieve labour market success. This would result in a positive correlation between college proximity and the unobserved part of the earnings equation, yielding an invalid instrument. A second possibility Card discusses is that the presence of a college could be associated with higher quality local primary and secondary schools. As school quality effects wages\footnote{See Card and Krueger 1992.} the omission of this variable would lead to an error in wages that is correlated to college proximity. One last possibility Card mentions is that if there are only imperfect indicators for place of residence, and if those who grew up with a college near by also live in higher-wage areas, then college proximity would be correlated with an unobserved geographic wage premium. 

In general one cannot test if an instrument is exogenous, however, in this case Card proposed a strategy. Recall, the interpretation of college proximity as a factor that lowers the cost of schooling suggests that the effect of college proximity should be largest for poor families. Due to this one can include college proximity directly in the earnings equation and then use an interaction between college proximity and an indicator variable for low parental education as an instrument for eduation. 

Table 6 presents the reduced form estimates and structural equation using this instrument. Specifically, low family background is defined as niether parent graduating from highschool. Then, the interaction with this indicator variable and proximty to a 4 year college is used to instrument education. The coefficients in columns 1 and 2 verify our suspision that the effects of living near college are larger for those with low family background. Moreover, in column 3 we can see the estimated return to schooling is about the same as our prior IV estimates. It is also notable that in this regression the coefficient on college proximity itself is small and not signficantly different from zero. Altogether, these results provide no evidence against the assumption that college proximity does not directly effect earnings and gives us more confidence that it is a valid instrument.

# Conclusion

This short paper recreated the main analysis done by Card in his 1993 working paper, where he uses college proximity as an instrument for education. The results from Tables 2-5 suggest that conventional OLS estimates of the return to schooling are downward-biased. In particular, the IV estimates suggest the return to schooling is about 12.2 to 13.6 percent whereas the OLS estimates suggest it is about 7.3 percent. 

Like Card's findings, these results are robust to small changes in specification. Specifically, the inclusion of KWW scores and using proximity to 2-year and 4-year colleges as instruments did not signficantly change the findings. Despite this, for college proximity to be a valid instrument we rely on the relatively strong assumption that college proximity has no direct effect on earnings. To test this I repeated the approach taken by Card. Recall, college proximity has a larger effect on the schooling decision for individuals with a poor family background. Therefore, an interaction between poor family background and college proximity can be used to instrument education in models where college proximity is assumed to have a direct effect on earnings. The results of this specification imply a return to schooling similar to the basic IV specifications, and the coefficient on college proximity was not significantly different from zero. 

As Card concluded, while these estimates are not very precise they all imply that the return to schooling is relatively high for students with poor family background. As Card notes, these results also go against the conventional wisdom that individuals with higher education would earn above-average income regardless of education.


# References
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\noindent

Card, D., & Krueger, A. B. (1992). Does school quality matter? Return to education and the characteristics of public schools in the United States. \textit{Journal of political Economy}, 100(1), 1-40.

Card, D. (1993). Using geographic variation in college proximity to estimate the return to schooling. \textit{NBER working paper}, (w4483).

Card, D. (1999). The causal effect of education on earnings. \textit{Handbook of labor economics}, 3, 1801-1863

\newpage 

```{r, eval=TRUE}
card <- wooldridge::card

#For geographic proximity to be a valid instrument it must be relevant and exogenous. That is, it must effect the schooling decision but have no direct effect on earnings. 

#To be a valid instrument for education, the variable must effect the schooling decision but have no direct effect on earnings.

# Ones proximity variation in the distance to college as a source of exogenous variation in education choices.
```



```{r,eval=TRUE}
card <- card %>%
  # Experience squared / 100
  mutate(exper2 = (exper^2)/100,
  # parent education, assigned mean if na, and indicator
  fatheduc_new = ifelse(is.na(fatheduc),mean(card$fatheduc,na.rm=T),fatheduc),
  fatheduc_missing = ifelse(is.na(fatheduc),1,0),
  motheduc_new = ifelse(is.na(motheduc),mean(card$motheduc,na.rm=T),motheduc),
  motheduc_missing = ifelse(is.na(motheduc),1,0),
  
  #age^2 /100 as instrument for experience
  age2=(age^2)/100
  )
```

```{r,eval=TRUE}
controls <- "educ + exper + exper2 + black + south + smsa"
regs_66 <- "reg661+reg662+reg663+reg664+reg665+reg666+reg667+reg668+smsa66"
fam_educ <- "motheduc_new+motheduc_missing+fatheduc_new+fatheduc_missing"
fam_educ_sq <- "motheduc_new+motheduc_missing+fatheduc_new+fatheduc_missing+motheduc_new:fatheduc_new+motheduc_new:fatheduc_missing + motheduc_new:motheduc_missing+motheduc_missing:fatheduc_new+motheduc_missing:fatheduc_missing+fatheduc_missing:fatheduc_new"
fam_structure <- "momdad14+sinmom14"

# OLS
ols <- list(
  "(1)"=lm(formula(paste0("lwage ~",controls)) ,data=card),
  "(2)"=lm(formula(paste0("lwage ~",controls,"+",regs_66)),data=card),
  "(3)"=lm(formula(paste0("lwage ~",controls,"+",regs_66,"+",fam_educ)),
           data=card),
  "(4)"=lm(formula(paste0("lwage ~",controls,"+",regs_66,"+",fam_educ_sq)),
           data=card),
  "(5)"=lm(formula(paste0("lwage ~",controls,"+",regs_66,"+",fam_educ_sq,"+",
           fam_structure)),data=card)
  )
```


```{r,eval=TRUE}
# table
modelsummary(ols,
             coef_map = c("educ" = "Education", 
                          "exper"="Experience",
                          "exper2"="Experience Squared / 100",
                          "black"="Race Indicator",
                          "south"="Live in South",
                          "smsa"="Live in SMSA"),
             #robust se
             statistic_override = sandwich,
             stars = F,
             gof_omit = "AIC|BIC|Log.Lik.|F",
             title = "OLS Regressions for Log Hourly Earnings",
             notes = "Notes: Column 2 includes indicators for regions. Column 3 includes mother and father education levels. Column 4 includes interactions between these education terms. Column 5 includes indicators for family structure. "
             )%>% 
  kable_classic_2()

```





```{r, eval=T}
controls <- "exper + exper2 + black + south + smsa"

tsls <- list(
  "Education" = feols(formula(paste0("educ ~ nearc4 + ", controls,"+",regs_66)), data=card, se="hetero"),
  "Education (2)" = feols(formula(paste0("educ ~ nearc4 + ", controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq)), data=card ,se="hetero"),
  "Earnings" = feols(formula(paste0("lwage ~ nearc4 +", controls,"+",regs_66)),data=card,se="hetero"),
  "Earnings (2)" = feols(formula(paste0("lwage ~ nearc4 +", controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq)),data=card,se="hetero"),
  "IV " = feols(formula(paste0("lwage ~ ",controls,"+",regs_66,"| educ ~ nearc4")),data=card,se="hetero"),
  "IV (2)" = feols(formula(paste0("lwage ~ ",controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"| educ ~ nearc4")),data=card, se="hetero")
  )
```  

```{r,eval=T}
# table
modelsummary(tsls,
             coef_map = c("fit_educ" = "Education (Fit)", 
                          "nearc4"="Near College"),
             stars = F,
             gof_omit = "AIC|BIC|Log.Lik.|F|Std. errors|R2 Within|R2 Pseudo|R2|R2 Adj.",
             title = "IV Regressions for Log Hourly Earnings",
             notes = "Notes: Columns 1, 3, and 5 include controls for experience and indicators for race and regions. Columns 2, 4, and 6  includes controls for parents education and family structure. Age used as instrument for experience."
             )%>% 
  kable_classic_2()

```

```{r, eval=T}
controls <- "black + south + smsa"

tsls_age <- list(
  "Education" = feols(formula(paste0("educ ~ nearc4 + ", controls,"+",regs_66,"|exper + exper2 ~ age + age2")), data=card, se="hetero"),
  "Education (2)" = feols(formula(paste0("educ ~ nearc4 + ", controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"|exper + exper2 ~ age + age2")), data=card ,se="hetero"),
  "Earnings" = feols(formula(paste0("lwage ~ nearc4 +", controls, "+", regs_66, "|exper + exper2 ~ age + age2")),data=card,se="hetero"),
  "Earnings (2)" = feols(formula(paste0("lwage ~ nearc4 +", controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"|exper + exper2 ~ age + age2")),data=card,se="hetero"),
  "IV" = feols(formula(paste0("lwage ~ ",controls,"+",regs_66,"| educ + exper +exper2~ nearc4+age+age2")),data=card,se="hetero"),
  "IV (2)" = feols(formula(paste0("lwage ~ ",controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"| educ +exper+exper2~ nearc4+age+age2")),data=card, se="hetero")
  )
```  

```{r,eval=T}
# table
modelsummary(tsls_age,
             coef_map = c("fit_educ" = "Education (Fit)", 
                          "nearc4"="Near College"),
             stars = F,
             gof_omit = "AIC|BIC|Log.Lik.|F|Std. errors|R2 Within|R2 Pseudo|R2|R2 Adj.",
             title = "IV Regressions for Log Hourly Earnings, Age as Instrument for Experience",
             notes = "Notes: Columns 1, 3, and 5 include controls for experience and indicators for race and regions. Columns 2, 4, and 6  includes controls for parents education and family structure. Age used as instrument for experience."
             )%>% 
  kable_classic_2()

```


```{r, eval=T}
controls <- "black + south + smsa"

alt_ols <- list(
  "Basic OLS Estimate" = feols(formula(paste0("lwage ~ educ +",controls,"+",regs_66,"+",fam_educ_sq,"+", fam_structure,"|exper+exper2~age+age2")),data=card),
  "KWW" = feols(formula(paste0("lwage ~ educ +",controls,"+ KWW  +",regs_66,"+",fam_educ_sq,"+",fam_structure,"|exper+exper2~age+age2")),data=card),
  "KWW with Instrument" = feols(formula(paste0("lwage ~ educ + KWW + ",controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"| KWW+ exper+exper2~ IQ+age+age2")),data=card)
)


```  

```{r,eval=T}
# table
modelsummary(alt_ols,
             coef_map = c("educ" = "Education"), 
                          #"KWW"="KWW Score",
                          #"fit_KWW"="KWW Score (Fit)"),
             #robust se
             statistic_override = sandwich,
             stars = F,
             gof_omit = "AIC|BIC|Log.Lik.|F|Std. errors|R2 Within|R2 Pseudo|R2|R2 Adj.",
             title = "Alternate OLS Specifications for Log Hourly Earnings",
            notes = "Notes: Column 1 includes controls for experience, race, regions, parental education and family structure. Age is used as an instrument for experience. Column 2 includes KWW score. Column 3 uses IQ as an instrument for KWW."
             )%>% kable_classic_2()

```
```{r,eval=T}
alt_tsls <- list(
  "Basic IV Estimate" = feols(formula(paste0("lwage ~ ",controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"| educ+exper+exper2 ~ nearc4+age+age2")),data=card, se="hetero"),
    "KWW" = feols(formula(paste0("lwage ~ KWW+",controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq," |educ+exper+exper2 ~ nearc4+age+age2 ")),data=card, se="hetero"),
    "KWW with Instrument" = feols(formula(paste0("lwage ~ ",controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"| educ + KWW +exper+exper2~ nearc4 + IQ+age+age2")),data=card, se="hetero"),
    "Near 2-yr & 4-yr College" = feols(formula(paste0("lwage ~ ",controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"| exper+exper2+educ ~ age+age2+nearc2 + nearc4")),data=card, se="hetero")
  )
```

```{r,eval=T}
# table
modelsummary(alt_tsls,
             coef_map = c("fit_educ" = "Education (Fit)"), 
                        #  "KWW" = "KWW Score",
                         # "fit_KWW" = "KWW Score (Fit)"),
             stars = F,
             gof_omit = "AIC|BIC|Log.Lik.|F|Std. errors|R2 Within|R2 Pseudo|R2|R2 Adj.",
             title = "Alternate IV Specfications for Log Hourly Earnings",
             notes = "Notes: Column 1 includes controls for experience, race, regions, parental education and family structure. Age is used as an instrument for experience. Column 2 includes KWW score. Column 3 uses IQ as an instrument for KWW score. Column 4 does not include KWW score, but uses proximity to 2-year and 4-year colleges as instruments for education."
             )%>% 
  kable_classic_2()

```


```{r,eval=T}
card <- card %>%
  mutate(
    low_par_ed = ifelse(motheduc_new<12 & fatheduc_new<12,1,0)
  )
```


```{r,eval=T}
controls <- "black + south + smsa"

tsls2 <- list(
  "Education" = feols(formula(paste0("educ ~ nearc4*low_par_ed+", controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"|exper + exper2 ~ age + age2")), data=card ,se="hetero"),
  "Earnings" = feols(formula(paste0("lwage ~ nearc4*low_par_ed +", controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"|exper + exper2 ~ age + age2")),data=card,se="hetero"),
    "Structural Model" = feols(formula(paste0("lwage ~ low_par_ed+nearc4+", controls,"+",regs_66,"+",fam_structure,"+",fam_educ_sq,"| exper + exper2 + educ ~ age + age2 + nearc4:low_par_ed")),data=card,se="hetero")
)
```



```{r,eval=T}
modelsummary(tsls2,
             coef_map = c("fit_educ" = "Education (Fit)", 
                          "nearc4"="Near College",
                          "nearc4:low_par_ed"="Near College * Low Parental Education"),
             stars = F,
             gof_omit = "AIC|BIC|Log.Lik.|F|Std. errors|R2 Within|R2 Pseudo|R2|R2 Adj.",
             title = "Interaction of Parent Education and College Proximity as Instrument for Education",
             notes = "Notes: All columns includes controls for experience, race, regions, parental education and family structure. Age is used as an instrument for experience. The interaction term between low parental education and near 4-year college is used as an instrument for education."
             )%>% 
  kable_classic_2()
```



